Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-05-13 paper.qmd
Put the intro image to first page
Updated Figure 3 text

Non-linear dimension reduction (NLDR) is popular for making a convenient low-dimensional (\kD{}) representation of high-dimensional (\pD{}) data ($k < p$). Recently developed methods include t-distributed stochastic neighbor embedding (tSNE) [@laurens2008], uniform manifold approximation and projection (UMAP) [@leland2018], potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm [@moon2019], large-scale dimensionality reduction Using triplets (TriMAP) [@amid2019], and pairwise controlled manifold approximation (PaCMAP) [@yingfan2021]. However, the representation generated can vary dramatically from method to method, and with different choices of parameters or random seeds made using the same method (@fig-NLDR-variety). The dilemma for the analyst is then, **which representation to use**. The choice might result in different procedures used in the downstream analysis, or different inferential conclusions. The research described here provides new visual tools to aid with this decision.

@@

@@

hex_grid_temp45 <- hex_grid_temp |> 
  filter(hex_poly_id == 45)

@@

a1 <- tibble(x = all_centroids_df_temp$c_x[4],
             xend = all_centroids_df_temp$c_x[5],
             y = all_centroids_df_temp$c_y[21],
             yend = all_centroids_df_temp$c_y[21],
             label = expression(a[1]))
a2 <- tibble(x = all_centroids_df_temp$c_x[25],
             xend = all_centroids_df_temp$c_x[25],
             y = all_centroids_df_temp$c_y[25],
             yend = all_centroids_df_temp$c_y[33],
             label = expression(a[2]))
l <- tibble(x = hex_grid_temp45$x[2],
            xend = hex_grid_temp45$x[3],
            y = hex_grid_temp45$y[2],
            yend = hex_grid_temp45$y[3],
            label = expression(l))

@@

             hjust=-0.1, vjust=-0.3) + 

@@

             hjust=1.1, vjust=-0.3) + 

@@

            hjust=-0.1, vjust=1.2) + 

@@

             label = expression(r[1]), color = "black") +

@@

             label = expression(r[2]), color = "black") +

@@

             vjust = 1.2) +

@@

             color = "black", hjust=-0.2) +

@@

>>>>

Non-linear dimension reduction (NLDR) is popular for making a convenient low-dimensional (\kD{}) representation of high-dimensional (\pD{}) data ($k < p$). Recently developed methods include t-distributed stochastic neighbor embedding (tSNE) [@laurens2008], uniform manifold approximation and projection (UMAP) [@leland2018], potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm [@moon2019], large-scale dimensionality reduction Using triplets (TriMAP) [@amid2019], and pairwise controlled manifold approximation (PaCMAP) [@yingfan2021]. However, the representation generated can vary dramatically from method to method, and with different choices of parameters or random seeds made using the same method (@fig-NLDR-variety).

@@

The dilemma for the analyst is then, **which representation to use**. The choice might result in different procedures used in the downstream analysis, or different inferential conclusions. The research described here provides new visual tools to aid with this decision.

@@

hex_grid_temp40 <- hex_grid_temp |> 
  filter(hex_poly_id == 40)

@@

a1 <- tibble(x = all_centroids_df_temp$c_x[4],
             xend = all_centroids_df_temp$c_x[5],
             y = all_centroids_df_temp$c_y[21],
             yend = all_centroids_df_temp$c_y[21],
             label = expression(a[1]))
a2 <- tibble(x = all_centroids_df_temp$c_x[25],
             xend = all_centroids_df_temp$c_x[25],
             y = all_centroids_df_temp$c_y[25],
             yend = all_centroids_df_temp$c_y[33],
             label = expression(a[2]))
l <- tibble(x = hex_grid_temp40$x[2],
            xend = hex_grid_temp40$x[3],
            y = hex_grid_temp40$y[2],
            yend = hex_grid_temp40$y[3],
            label = expression(l))

@@

             hjust=-0.1, vjust=-0.3, size = 8) + 

@@

             hjust=1.1, vjust=-0.3, size = 8) + 

@@

            hjust=-0.1, vjust=1.2, size = 8) + 

@@

             label = expression(r[1]), color = "black", size = 8) +

@@

             label = expression(r[2]), color = "black", size = 8) +

@@

             vjust = 1.2, size = 8) +

@@

             color = "black", hjust=-0.2, size = 8) +

@@

      color = "black", size = 8) +
